Statically Typed String Sanitation Inside a Python

### 2014-08-18 sanitation-psp14/sanitation-psp14-short.tex

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications \cite{OWASP}. They arise because web applications, at their boundaries, control external systems by issuing  commands represented using  strings. For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings. When these commands must include data derived from user input, care must be taken to ensure that the user cannot construct input that subverts the intended command. For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability: 
\begin{lstlisting}[numbers=none]
'SELECT * FROM users WHERE name="' + name + '"'
\end{lstlisting}
If a malicious user enters the name \lstinline{'"; DROP TABLE users --'}, the entire database could be erased.

To avoid this problem, the program must \emph{sanitize} user input. For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes (and existing backslashes)  with a backslash, which SQL treats safely. Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging. Note that this function is not idempotent, so it should only be called once.

We observe that most such sanitization techniques can be understood in terms of \emph{regular languages} \cite{cinderella}. For example, \verb|name| should be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*! -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped. This concrete pattern syntax can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction. We will work with this ``core'' for simplicity.

In this paper, we present a static type system that tracks the regular language a string belongs to. For example, the output of \verb|sanitize| will be in the regular language described by the regular expression above. We write this as $\mathcal{L}\{r\}$. We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}, including \emph{replacement} of substrings matching a given regular expression. These make it possible to implement sanitation protocols like the one just described in an essentially conventional manner. The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type. Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e. these functions need not be trusted). These guarantees require run-time checks only when going from less precise to more precise types (e.g. at the edges of the system, where user input has not yet been validated). %implementing and statically checking input sanitation techniques. Our solution suggests a more general approach to the integration of security concerns into programming language design. This approach is characterized by \emph{composable} type system extensions which \emph{complement} existing and well-understood solutions with compile-time checks.

@@

\begin{figure}[h]
\begin{grammar}
<$r$> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

@@

@@

<$e$> ::= 
      $x$ \hfill source terms \alt 
      $\lambda x . e$ \alt
      $e~(e)$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{e}{e}$ \alt
      $\sisubst{r}{e}{e}$ \alt
      $\rcoerce{r}{e}$ \alt
      $\rcheck{r}{e}{x.e}{e}$
\caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
\end{grammar}
\end{figure}

In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$. This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Sec. \ref{translation}.

The syntax of $\lambda_{RS}$ is given in Figure \ref{fig:lamrs}, its static semantics are specified in Figure \ref{fig:etyping} and its evaluation semantics are specified in Figure \ref{fig:eeval}.

The system has regular expression types $\stringin{r}$ whre $r$ is a regular expression. 
Expressions of this type evaluate to string literals in the language described by $r$.
Operations on expressions of type $\stringin{r}$ preserve this property.

@@

The $\lambdas$ system extends the simply-typed lambda calculus with regular
expression types; an explanation of significant typing rule follows:
\begin{itemize}
  \item Rule S-T-Concat states that the type of concatenated strings is 
    obtained by concatenating the regular expressions for each string.
  \item Rule S-T-Replace defines the type of $\textsf{rreplace}$. The expression $\textsf{rreplace}$
    evaluates to the result of substituting every string matching $r$ in $s_1$
    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
    function, which is defined later in this section.
    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
  \item Rule S-T-SafeCoerce allows only safe coercions between string types by
    expoiting the decidability of language inclusion.
  \item The check rules allow casts between strings that cannot be guaranteed at compile time, but might pass a runtime check.
\end{itemize}

\newpage

@@

@@

$\fbox{\inferrule{}{\sctx \vdash \trden{e} = \iota}}$
\begin{mathpar}

@@

\begin{figure}[h]\begin{lstlisting}
@fn
def sanitize(s : string_in[r'.*']):
  return (s.replace(r'"', '&quot;') 
           .replace(r'<', '&lt;')
           .replace(r'>', '&gt;'))

@@

def main(db):
  input = sanitize(user_input())
  results = db.execute(results_query(input))
  return results_div(input) + format(results)
\end{lstlisting}
\vspace{-10px}
\caption{Regular string types in atlang, a library that enables static type checking for Python. }   \label{examplefig}

\end{figure}

@@

  def check_Coerce(self, ctx, node, other_t):
    # coercions can only be defined between 
    # types with the same type constructor, 
    if rx_sublang(other_t.idx, self.idx):
      return other_t
    else: raise atlang.TypeError("...", node)
\end{lstlisting}
\vspace{-10px}
\caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
\end{figure}

@@

\subsection{Related Work and Alternative Approaches}

>>>>

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications \cite{OWASP}. They arise because web applications, at their boundaries, control external systems using commands represented as  strings. For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings. When these commands include data derived from user input, care must be taken to ensure that the user cannot  subverts the intended command by carefully crafting the data they send. For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability: 
\begin{lstlisting}[numbers=none]
'SELECT * FROM users WHERE name="' + name + '"'
\end{lstlisting}
If a malicious user enters the name \lstinline{'"; DROP TABLE users --'}, the entire database could be erased.

To avoid this problem, the program must \emph{sanitize} user input. For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes double quotes and existing backslashes   with a backslash, which SQL treats safely. Guaranteeing that user input has already been sanitized like this before it is used to construct a command is challenging. Note that this function is not idempotent, so it should only be called once.

We observe that many such sanitization techniques can be understood using \emph{regular languages} \cite{cinderella}. For example, \verb|name| must be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*! -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped. This concrete syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction. We will work with this ``core'' for simplicity.

In this paper, we present a static type system that tracks the regular language a string belongs to. For example, the output of \verb|sanitize| will be a string in the regular language described by the regular expression above (we describe such a regular language, following convention, as $\mathcal{L}\{r\}$). By leveraging closure and decidability properties of regular languages, the language of a string is tracked through uses of a number of operations, including \emph{replacement} of substrings matching a given regular expression. This makes it simple to implement sanitation procedures like the one just described in an essentially conventional manner. The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type. Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e. these functions need not be trusted). These guarantees require run-time checks only when going from less precise to more precise types (e.g. at the edges of the system, where user input has not yet been validated). %implementing and statically checking input sanitation techniques. Our solution suggests a more general approach to the integration of security concerns into programming language design. This approach is characterized by \emph{composable} type system extensions which \emph{complement} existing and well-understood solutions with compile-time checks.

@@

\begin{figure}[t]
\begin{grammar}
<$r$> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

@@

\begin{figure}[t]
\begin{grammar}

@@

<$e$> ::= 
      $x$ \hfill source terms \alt 
      $\lambda x . e$ \alt
      $e(e)$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{e}{e}$ \alt
      $\sisubst{r}{e}{e}$ \alt
      $\rcoerce{r}{e}$ \alt
      $\rcheck{r}{e}{x.e}{e}$
\caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
\end{grammar}
\end{figure}

In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$. The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:lamrs}, its static semantics in Figure \ref{fig:etyping} and its evaluation semantics in Figure \ref{fig:eeval}. This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Sec. \ref{translation}.

There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$. Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}. Regular string types are of the form $\stringin{r}$, where $r$ is a regular expression. Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e. $s \in \Sigma^\star$, defined in the usual way). The rule \textsc{S-T-Stringin-I} ensures that $s \in \lang{r}$. The remaining operations on terms of type $\stringin{r}$ preserve this property. %The remaining operators operate on regular strings.

An explanation of significant typing rules follows:
\begin{itemize}
  \item \textsc{S-T-Concat}: the type of concatenated regular strings is  
    obtained by concatenating the regular expressions.
  \item \textsc{S-T-Replace}: \todo{fix this description}the term $\textsf{rreplace}$
    evaluates to the result of substituting every string matching $r$ in $s_1$
    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
    function, which is defined later in this section.
    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
  \item \textsc{S-T-SafeCoerce}: \todo{fix this description} allows only safe coercions between string types by
    expoiting the decidability of language inclusion.
  \item \textsc{S-T-Check}: \todo{fix this description}The check rules allow casts between strings that cannot be guaranteed at compile time, but might pass a runtime check.
  \item \todo{strcase}
\end{itemize}

@@

@@

The $\lambdas$ system extends the simply-typed lambda calculus with regular
expression types

%\newpage

@@

$\fbox{\inferrule{}{\sctx \vdash \trden{e} = \iota}}$
\begin{mathpar}
\inferrule[Tr-Var]
{ }
{\sctx \vdash \trden{x} = x}

@@

\section{Implementation in {Atlang}}\label{atlang}
In the previous section, we specified a type system and a translation semantics to a language containing only strings and regular expressions. In this section, we take Python to be such a target language. Python does not have a static type system, however, so to implement these semantics, we will leverage \verb|atlang|, an extensible type system for Python (being developed by the authors). By using \verb|atlang|, which builds on Python's quotations and reflection facilities, we can implement these semantics as a library, rather than as a new dialect of Python.

\begin{figure}[h]\begin{lstlisting}
from atlib import fn, string_in

@fn
def sanitize(s : string_in[r'.*']):
  return (s.replace(r'"', '&quot;') 
           .replace(r'<', '&lt;')
           .replace(r'>', '&gt;'))

@@

@fn
def main(db):
  input = sanitize(user_input())
  results = db.execute(results_query(input))
  return results_div(input) + format(results)
\end{lstlisting}
\vspace{-10px}
\caption{Regular string types in atlang, a library that enables static type checking for Python. }\label{fig:atexample}
\end{figure}

@@

  def check_Coerce(self, ctx, node, other_t):
    # coercions can only be defined between 
    # types with the same type constructor, 
    if rx_sublang(other_t.idx, self.idx):
      return other_t
    else: raise atlang.TypeError("...", node)
\end{lstlisting}
\vspace{-10px}
\caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
\label{fig:atimpl}\end{figure}

\subsection{Example Usage}

Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|string_in|, both of which we have included in \verb|atlib|, the standard library for \verb|atlang|. The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|atlang|. The function \verb|sanitize| on lines 3-7 specifies one argument, \verb|s|, of type \lstinline{string_in[r'.*']}.\todo{fix sizes}

The sanitize function takes an arbitrary string and returns a string without
double quotes or left and right brackets. In this example, we use HTML
escape sequences.

@@

\section{Related Work and Alternative Approaches}\label{related}
